
---
title: "Flights at Bay"
author: "Akshat Bhushan Sinha"
date: "October 2026"
output:
  ioslides_presentation:
    widescreen: true
    smaller: false
---

<style>
body, p, li, table {
  font-family: "Times New Roman", Times, serif !important;
}
slides > slide {
  background: #F4F8F5 !important;
  color: #18352B !important;
}
h1, h2, h3 {
  font-family: "Times New Roman", Times, serif !important;
  color: #064E3B !important;
}
strong {
  color: #087F5B;
}
.title-slide {
  background: #064E3B !important;
  color: #FFFFFF !important;
}
.title-slide hgroup h1,
.title-slide hgroup p {
  color: #FFFFFF !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

library(nycflights13)
library(dplyr)
library(ggplot2)

flight_data <- nycflights13::flights |>
  filter(
    !is.na(dep_delay),
    !is.na(hour),
    !is.na(distance)
  ) |>
  mutate(
    delayed = dep_delay > 15
  )

airport_summary <- flight_data |>
  group_by(origin) |>
  summarise(
    total_flights = n(),
    delay_rate = mean(delayed) * 100,
    .groups = "drop"
  )

hour_summary <- flight_data |>
  group_by(hour) |>
  summarise(
    delay_rate = mean(delayed) * 100,
    flights = n(),
    .groups = "drop"
  )
```

## Flights at Bay

### Keep flight delays at bay.

**AVIATION INTELLIGENCE · DELAY ANALYTICS**

An interactive R Shiny application exploring historical departure delays at New York's three major airports.

**The idea:** turn flight records into an accessible tool for exploring delay patterns and estimated risk.

## The problem

### Delays disrupt journeys

- Passengers need better ways to understand historical delay patterns.
- Departure delays may vary by airport and scheduled departure hour.
- Historical data can reveal patterns worth investigating.

**Our objective**

Build an interactive dashboard that lets users explore flight scenarios and understand the limitations of delay estimates.

## Data and methodology

### From flight records to an estimate

```{r data-summary}
data.frame(
  Measure = c(
    "Historical flight records",
    "Departure airports",
    "Dataset year",
    "Delay threshold"
  ),
  Value = c(
    format(nrow(flight_data), big.mark = ","),
    "JFK, LGA, EWR",
    "2013",
    "More than 15 minutes"
  )
)
```

**Data source:** `nycflights13`, a dataset of New York-area flights from 2013.

**Application inputs**
- Departure airport
- Scheduled departure hour
- Flight distance in miles

**Model:** logistic regression estimates the probability of a departure delay exceeding 15 minutes using the selected airport, hour and distance.

## Explore the patterns

### Historical departure delays

```{r delay-chart, fig.width=9, fig.height=4}
ggplot(
  airport_summary,
  aes(
    x = reorder(origin, delay_rate),
    y = delay_rate
  )
) +
  geom_col(
    fill = "#0B8F68",
    width = 0.65
  ) +
  geom_text(
    aes(label = sprintf("%.1f%%", delay_rate)),
    hjust = -0.15,
    color = "#18352B",
    size = 4
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.18))
  ) +
  labs(
    x = NULL,
    y = "Flights delayed over 15 minutes",
    title = "Departure delay rate by airport"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    text = element_text(
      family = "serif",
      color = "#18352B"
    ),
    plot.title = element_text(
      color = "#064E3B",
      face = "bold"
    ),
    plot.background = element_rect(
      fill = "#F4F8F5",
      color = NA
    ),
    panel.background = element_rect(
      fill = "#F4F8F5",
      color = NA
    ),
    panel.grid.major = element_line(
      color = "#D4E4DA"
    ),
    panel.grid.minor = element_blank()
  )
```

**In the Shiny app**, users can select an airport, adjust departure hour and distance, and run the model to explore an estimated delay probability.

## Findings and limitations

### What can we conclude?

- Historical delay rates can be compared across airports.
- Departure-hour patterns can be explored visually.
- Interactive inputs make the model accessible to non-technical users.

**Limitations**
- The dataset represents 2013, not live operations.
- Historical associations do not prove causation.
- The model is educational and has not been validated for operational forecasting.

### Flights at Bay

**Understand delays. Make informed decisions.**

Built with R, Shiny, `nycflights13`, `dplyr` and `ggplot2`.
